---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "90%",
  fig.align = "center",
  dpi = 300
)
```

# Bias Correction by Zero-Inflated Vine Copulas <img src="man/figures/logo.svg" align="right" height="139" alt="" />

<!-- badges: start -->
<!-- badges: end -->

**Vine Copula Bias Correction (VBC)** is a multivariate bias correction
methodology grounded in vine copula theory. The method generalizes marginal
distributions, copulas, and vine copula densities to accommodate mixture
distributions, and extends the Rosenblatt transform to handle non-continuous
pseudo-observations. This allows VBC to model dependencies among 
**heavy-tailed, zero-inflated, and continuous** climate variables at high
temporal resolution, providing more accurate and robust corrections.

A key feature of VBC is its **interpretability**: users can transparently 
control the correction process and assess the outcomes using built-in
diagnostics. For the full methodological details, please see our [paper](https://doi.org/10.1093/jrsssc/qlaf044).

## Installation

``` r
# without vignette
remotes::install_github("henrifnk/VBC")
# with vignette
remotes::install_github("henrifnk/VBC", build_vignettes = TRUE)
```

## Quick Start

We aim to correct CRCM5 climate data for Munich, Germany, for the year 2010.
The data is available in the package as `climate`.
We will use the `vbc` function to correct the data. 


```{r lib_n_data, message=FALSE}
library(data.table)
library(ggplot2)
library(knitr)
library(patchwork)
#library(VBC)

data("climate")

climate_2010 = lapply(climate, function(data) data[year(time) == 2010, ])
```

```{r, include=FALSE, echo=FALSE}
devtools::load_all()
```


Climate data are available in 3-hourly resolution for the variables temperature, precipitation, dew point temperature, radiation and wind speed.
The high temporal resolution causes inflation in the variables radiation and precipitation. 
Visually, check the shape of the marginal distributions of the model data and the reference data using `plot_tails`.
The distance between the two marginal distributions can be quantified using the Wasserstein distance. 

```{r zeroinflation, warning=FALSE, message=FALSE}
plot_tails(climate_2010$mp, "pr", scale_d = 0.1, mult = 4, xmin = 0) +
  labs(x = "", title = "Marginal distribution of uncorrected model data") +
  theme(legend.position = "none") +
  plot_tails(climate_2010$rp, "pr", scale_d = 1, mult = 4, xmin = 0) +
  labs(x = "precipitation (mm/h)",  
       title = "Marginal targeted distribution of reference data") +
  scale_y_continuous(name = "") +
  plot_layout(ncol = 1) 

wd_pre = calc_wasserstein(climate_2010$mp[, "pr"], climate_2010$rp[, "pr"])
wd_pre
```

For the correction, we need to specify the type of margins and their limits.
`"zi"` defines a univariate margins and `"c"` a strictly continuous margin.
`xmin` specifies the lower bound of the margins.
For the vine copula modeling, we use the *TLL* family set with no truncation
on the vine.

```{r modelling, echo = FALSE}
margins_controls <- list(xmin = c(NaN, 0, NaN, 0, 0),
                         type = c("c", "zi", "c", "zi", "c"))
mp_vbc = vbc(mp = climate_2010$mp[, -"time"], mc = climate_2010$mp[, -"time"],
             rc = climate_2010$rp[, -"time"],
             margins_controls = margins_controls, family_set = "tll",
             trunc_lvl = Inf)    
```

We can then visually and quantitatively asses the correction in `mp_vbc` by
comparing the corrected data and the reference data by plotting the tails and
calculating the Wasserstein distances.
The results can be compared to those above.


```{r zi-correction, warning=FALSE, message=FALSE, fig.height=2.5}
plot_tails(round(mp_vbc, 3), "pr", scale_d = 1, mult = 3, xmin = 0) +
  labs(x = "precipitation (mm/h)",
       title = "Marginal distribution of corrected model data")
```

Visibly, the correction shortens the heavy tail and increases the density inflation at zero.
This is also reflected in the Wasserstein distance.

```{r}
wd_post = calc_wasserstein(climate_2010$rp[, "pr"], mp_vbc[, "pr"])

kable(data.frame("Wasserstein_Uncorrected" = wd_pre,
                 "Wasserstein_Corrected" = wd_post,
                 "Improvement" = wd_pre - wd_post,
                 "Improvement_in_Perc" = (wd_pre - wd_post) / wd_pre * 100),
      digits = 2)
```

Further we can quantify the multivariate improvement of our correction in 
terms of the Wasserstein distances.

```{r}
wd_mvd_post = calc_wasserstein(climate_2010$rp[, -"time"], mp_vbc)
wd_mvd_pre = calc_wasserstein(climate_2010$rp[, -"time"],
                              climate_2010$mp[, -"time"])
iprovement = wd_mvd_pre - wd_mvd_post

kable(data.frame("Wasserstein_Uncorrected" = wd_mvd_pre,
                 "Wasserstein_Corrected" = wd_mvd_post,
                 "Improvement" = iprovement,
                 "Improvement_in_Perc" = iprovement / wd_mvd_pre * 100),
      digits = 2)

```

# Citation

If you use `VBC` in a scientific publication, please
[cite](https://doi.org/10.1093/jrsssc/qlaf044) it as:

    Henri Funk, Ralf Ludwig, Helmut Küchenhoff, Thomas Nagler, Towards more realistic climate model outputs: a multivariate bias correction based on zero-inflated vine copulas, Journal of the Royal Statistical Society Series C: Applied Statistics, 2025;, qlaf044, https://doi.org/10.1093/jrsssc/qlaf044
    
BibTeX:

``` tex
@article{funk2025,
    author = {Funk, Henri and Ludwig, Ralf and Küchenhoff, Helmut and Nagler, Thomas},
    title = {Towards more realistic climate model outputs: a multivariate bias correction based on zero-inflated vine copulas},
    journal = {Journal of the Royal Statistical Society Series C: Applied Statistics},
    pages = {qlaf044},
    year = {2025},
    month = {08},
    issn = {0035-9254},
    doi = {10.1093/jrsssc/qlaf044},
    url = {https://doi.org/10.1093/jrsssc/qlaf044},
    eprint = {https://academic.oup.com/jrsssc/advance-article-pdf/doi/10.1093/jrsssc/qlaf044/64045829/qlaf044.pdf},
}
```
