---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "90%",
  fig.align = "center",
  dpi = 300
)
```

# Bias Correction by Zero-Inflated Vine Copulas <img src="man/figures/logo.svg" align="right" height="130" alt="" />

<!-- badges: start -->
<!-- badges: end -->

Vine Copula Bias Correction for partially zero-inflated margins (VBC) is a  multivariate bias correction methodology anchored in vine copula theory. For this method, we generalised the margins, the copula and the vine copula density to accommodate mixture distributions and propose an extension to the Rosenblatt transform that can handle non-continuous pseudo observations. The proposed method is tailored to model dependencies between heavy tailed, zero-inflated and continuous climate variables with high temporal resolution, ensuring precise corrections. Notably, our approach is designed for high interpretability, enabling control and assessment of the results. 

## Installation

You can install the development version of VBC from [GitHub](https://github.com/) with:

``` r
# without vignette
devtools::install_github("henrifnk/VBC")
# with vignette
devtools::install_github("henrifnk/VBC", build_vignettes = TRUE)
```

## Quickstart

We aim to correct CRCM5 climate data for Munich, Germany, for the year 2010.
The data is available in the package as `climate`.
We will use the `vbc` function to correct the data. 


```{r lib_n_data, message=FALSE}
library(data.table)
library(ggplot2)
library(knitr)
library(patchwork)
library(VBC)

data("climate")

climate_2010 = lapply(climate, function(data) data[year(time) == 2010, ])
```

Climate data are available in 3-hourly resolution for the variables temperature, precipitation, dew point temperature, radiation and wind speed.
The high temporal resolution causes inflation in the variables radiation and precipitation. 
Visually, check the shape of the marginal distributions of the model data and the reference data using `plot_tails`.
The distance between the two marginal distributions can be quantified using the Wasserstein distance. 

```{r zeroinflation, warning=FALSE, message=FALSE}
plot_tails(climate_2010$mp, "pr", scale_d = 0.2, offset = 3) +
  labs(x = "", title = "Marginal distribution of uncorrected model data") +
  xlim(c(-1, 15)) + theme(legend.position = "none") +
  plot_tails(climate_2010$rp, "pr", scale_d = 1.8, offset = 2) +
  labs(x = "precipitation (mm/h)",  
       title = "Marginal targeted distribution of reference data") +
  xlim(c(-1, 15)) + scale_y_continuous(name = "") +
  plot_layout(ncol = 1) 

wd_pre = calc_wasserstein(climate_2010$mp[, "pr"], climate_2010$rp[, "pr"])
wd_pre
```

For the correction, we need to specify the type of margins and their limits.
`"zi"` defines a univariate margins and `"c"` a stricktly continous margin.
`xmin` specifies the lower bound of the margins.
For the vine copula modelling, we use the *TLL* family set with no truncation
on the vine.

```{r modelling, echo = FALSE}
margins_controls <- list(xmin = c(NaN, 0, NaN, 0, 0),
                         type = c("c", "zi", "c", "zi", "c"))

mp_vbc = vbc(mp = climate_2010$mp[, -"time"], mc = climate_2010$mp[, -"time"],
             rc = climate_2010$rp[, -"time"],
             margins_controls = margins_controls, family_set = "tll",
             trunc_lvl = Inf)    
```

We can then visually and quantitatively asses the correction in `mp_vbc` by
comparing the corrected data and the reference data by plotting the tails and
calculating the Wasserstein distances.
The results can be compared to those above.


```{r zi-correction, warning=FALSE, message=FALSE, fig.height=2.5}
plot_tails(mp_vbc, "pr", scale_d = 1.8, offset = 20) +
  labs(x = "precipitation (mm/h)",
       title = "Marginal distribution of uncorrected model data")
```

```{r}
wd_post = calc_wasserstein(climate_2010$rp[, "pr"], mp_vbc[, "pr"])

kable(data.frame("Wasserstein_Uncorrected" = wd_pre,
                 "Wasserstein_Corrected" = wd_post,
                 "Improvement" = wd_pre - wd_post,
                 "Improvement_in_Perc" = (wd_pre - wd_post) / wd_pre * 100),
      digits = 2)
```

Further we can quantify the multivariate improvement of our correction in 
terms of the wasserstein distances.

```{r}
wd_mvd_post = calc_wasserstein(climate_2010$rp[, -"time"], mp_vbc)
wd_mvd_pre = calc_wasserstein(climate_2010$rp[, -"time"],
                              climate_2010$mp[, -"time"])
iprovement = wd_mvd_pre - wd_mvd_post

kable(data.frame("Wasserstein_Uncorrected" = wd_mvd_pre,
                 "Wasserstein_Corrected" = wd_mvd_post,
                 "Improvement" = iprovement,
                 "Improvement_in_Perc" = iprovement / wd_mvd_pre * 100),
      digits = 2)

```

# Citation

If you use `VBC` in a scientific publication, please
[cite](TBD) it as:

    APA Style...

BibTeX:

``` tex
@InProceedings{
TBD
}
```
